---
title: "Homework 2: The Big Short"
output: html_document
---
 
**This homework is due Friday February 26, 2016 at 5:00 PM. When complete, submit your code in the R Markdown file and the knitted HTML via GitHub.**

# Background 

This homework is motivated by circumstances surrounding the [financial crisis of 2007-2008](https://en.wikipedia.org/wiki/Financial_crisis_of_2007%E2%80%9308). We titled the homework _The Big Short_, after the book on the same topic that was also recently made into a movie.

Part of what caused the financial crisis was that the risk of certain [securities](https://en.wikipedia.org/wiki/Security_(finance)) sold by financial institutions were  underestimated. Specifically, 
the risk of mortgage-backed securities (MBS) and collateralized debt obligations (CDO), the price of which depends on homeowners making their monthly payments, was grossly underestimated. A combination of factors resulted in many more defaults than were expected. This resulted in a crash of the prices of these securities. As a consequence, banks lost so much money that they needed bailouts to avoid default.

Here we present a **very** simplified version of what happened with some of these securities. Hopefully it will help you understand how a wrong assumption about the statistical behavior of events can lead to substantial differences between what the model predicts and what actually happens. Specifically, we will see how using an independence assumption can result in misleading conclusions. Before we start with the specific application we ask you about a simple casino game.

# Problem 1

In the game of [roullete](https://en.wikipedia.org/wiki/Roulette)
you can bet on several things including black or red. On this bet, if you win, you double your earnings. How does the casino make money on this then? If you look at the [possibilities](http://www.math.uah.edu/stat/games/Roulette.png)
you realize that the chance of red or black are both slightly less than 1/2. There are two green spots, so the of landing on black (or red) is actually 18/38, or 9/19.


## Problem 1A

Let's make a quick sampling model for this simple version of roulette. You are going to bet a dollar each time you play and always bet on black. Make a box model for this process using the `sample` function. Write a function `get_outcome` that takes as an argument the number of times you play $N$ and returns your earnings $S_N$.

```{r}

roulette_wheel <- rep(c("Black","Red","Green"),c(18,18,2))

get_outcome <- function(N){
   sum(sample(ifelse(roulette_wheel=="Black", 1, -1), N, replace=TRUE))
}

```

## Problem 1B

Use Monte Carlo simulation to study the distribution of total earnings $S_N$ for $N=10,25,100,1000$. That is, study the distribution of earnings for different number of plays. What are the distributions of these two random variables? How do the expected values and standard errors change with $N$? Then do the same thing for the average winnings $S_N/N$. What result that you learned in class predicts this?

```{r}
B <- 10^5 #Replicates for simulation
mu <- -1/19 #sampling model average
########################################
# Total winnings


#Monte carlo simulation for total winnings
get_reps <- function(x) replicate(B,get_outcome(x))

outcome10 <- get_reps(10)
outcome25 <- get_reps(25)
outcome100 <- get_reps(100)
outcome1000 <-get_reps(1000)

#Placing in a list to make it easier to iterate over
sums <- list(outcome10,outcome25,outcome100,outcome1000)

#Visualising the distribution of total winnings over different values of N
lapply(sums,hist)

#From the histogram, the sums appear to be following a normal distribution
#so verifying with qqnorm
lapply(sums,qqnorm)

#Changes of Expected value of total winnings as N increases
lapply(sums,mean)

#SE for 1000 replicates, effect of changing N shown in the average
#winning section 
sqrt(mean((outcome1000-(mu*1000))^2))

#SE for 1000 reps from Math
(1000*2*sqrt(9/19*10/19))/sqrt(1000) 


####################################################
### Average winnings
get_avg_outcome <- function(N) get_outcome(N)/N

get_avg_reps <- function(x)replicate(B,get_avg_outcome(x))

avg_outcome10 <- get_avg_reps(10)
avg_outcome25 <- get_avg_reps(25)
avg_outcome100 <- get_avg_reps(100)
avg_outcome1000 <-get_avg_reps(1000)


means <- list(avg_outcome10,avg_outcome25,avg_outcome100,avg_outcome1000)

lapply(means,hist)
lapply(means,qqnorm)

#Changes of Expected value of average winnings as N increases
lapply(means, mean)

#Changes of SE of average winnings as N increases from estimate
lapply(means, function (x) sqrt(mean((x-mu)^2)))


############################
# Math section
binomial_prob <- function(a,b,pa) {
   pa*a + b*(1-pa)
}

#Expected value of betting on black from maths
binomial_prob(1,-1,9/19)

#SD
sigma <- 2*sqrt(9/19*10/19) 
sigma
#SE for win/loss
sigma/sqrt(1000)

```



Both of these random variables follow a normal distribution. 
As we increase the size of the samples, we can see a reduction in the standard error and the expected values converging on what we would expect to see mathematically.


## Problem 1C

What is the expected value of our sampling model? What is the standard deviation of our sampling model?

Expected value is -0.053 with a standard deviation of 0.999.

## Problem 1D

Use CLT to approximate the probability that the casino loses money when you play 25 times. Then use a Monte Carlo simulation to confirm.

```{r}

N<-25
mu <- binomial_prob(1,-1,9/19)*N    
sigma <- 2*sqrt(9/19*10/19)*N 

#z score of a positive profit for the player (loss for the casino)
z <- -sqrt(N)*mu/sigma

#prob(X>0)
1-pnorm(z)

# Monte carlo simulation
profit <- replicate(B,{
  X <- sample( c(-1,1), 25, replace=TRUE, prob=(c(10/19,9/19)) )
  mean(X)
})

#Casino loses when player generates a profit
#corresponds with what we predicted from the CLT
mean(profit>0)

```



## Problem 1E

In general, what is the probability that the casino loses money as a function of $N$? Make a plot for values ranging from 25 to 1,000. Why does the casino give you free drinks if you keep playing?

```{r}


l <- c(25:10000)
j <- sapply(l,function(N){
  mu <- binomial_prob(1,-1,9/19)*N    
  sigma <- 2*sqrt(9/19*10/19)*N 
  1-pnorm(-sqrt(N)*mu/sigma)
})

plot(j)

```

As the number of games played increase, the probability of the player winning tends downwards towards 0. The house always wins in the end.


# Problem 2 

You run a bank that has a history of identifying potential homeowners that can be trusted to make payments. In fact, historically, in a given year, only 2% of your customers default. You want to use stochastic models to get an idea of what interest rates you should charge to guarantee a profit this upcoming year. 

## Problem 2A

Your bank gives out 1,000 loans this year. Create a sampling model and use the function `sample` to simulate the number of foreclosure in a year with the information that 2% of customers default. Also suppose your bank loses $120,000 on each foreclosure. Run the simulation for one year and report your loss.

```{r}

p_default <- .02
default_cost <- 120000

foreclosure_sample <- function(){
   sample(c(1,0),10^3,prob=c(p_default,(1-p_default)),replace = TRUE)
}

foreclosures <- foreclosure_sample()

#Sanity check
prop.table(table(foreclosures))

loss <- sum(foreclosures)*default_cost
cat("Annual loss due to foreclosures: $",loss,sep="")

```

## Problem 2B

Note that the loss you will incur is a random variable. Use Monte Carlo simulation to estimate the distribution of this random variable. Use summaries and visualization to describe your potential losses to your board of trustees.

```{r}
losses <- replicate(B,mean(foreclosure_sample()))

hist(losses)
qqnorm(losses)
summary(losses)
```

## Problem 2C

The 1,000 loans you gave out were for $180,000. The way your bank can give out loans and not lose money is by charging an interest rate. If you charge an interest rate of, say, 2% you would earn $3,600 for each loan that doesn't foreclose. At what percentage should you set the interest rate so that your expected profit totals $100,000. Hint: Create a sampling model with expected value 100 so that when multiplied by the 1,000 loans you get an expectation of $100,000. Corroborate your answer with a Monte Carlo simulation.

Your solution here. 
```{r}

# unsuccessful loans cost the company 120000
# interest = (profit + unsuccessful loan cost)/(successful loan return)
total_profit <- 100000
n_loans <- 1000
loan_ammount <- 180000

get_interest <- function()(total_profit + (p_default*n_loans*default_cost))/((1-p_default)*n_loans*loan_ammount)


interest <- get_interest()

#simulation to check
get_profit <- function(){replicate(B, {
   X <- foreclosure_sample()
   sum(ifelse(X==1,-default_cost,interest*loan_ammount))
})}

profit = get_profit()
mean(profit)
  
  
```


## Problem 2D

In problem 2C, you were able to set a very low interest rate. Your customers will be very happy and you are expected to earn $100,000 in profits. However, that is just an expectation. Our profit is a random variable. If instead of a profit your bank loses money, your bank defaults. Under the conditions of Problem 2C, what is the probability that your profit is less than 0?

```{r}
mu <- 100000
sigma <- sqrt(mean((profit-mu)^2))

#probibility that profit is less than 0
pr <- pnorm((0-mu)/sigma/sqrt(n_loans))
pr


#monte carlo check
mean(profit<0)
```

## Problem 2E

Note that the probability of losing money is quite high. To what value would you have to raise interest rates in order to make the probability of losing money, and your bank and your job, as low as 0.001? What is the expected profit with this interest rate? Corroborate your answer with a Monte Carlo simulation.

Hint: Use the following short cut. If $p$ fraction of a box are $a$s and $(1-p)$ are $b$s, then the SD of the list is $\mid a-b \mid \sqrt{p(1-p)}$ 

Your solution here.

```{r}


sigma <- function(r) (180000*r + 120000)*sqrt(.02*.98)
mu <- function(r) (180000*.98 - 120000*.02)

se

qnorm(.999)

Profit <- replicate(B, {
   X <- foreclosure_sample()
   sum(ifelse(X==1,-120000,.02*180000))
})

mean(Profit)

```

## Problem 2F

Note that the Monte Carlo simulation gave a slightly higher probability than 0.001. What is a possible reason for this? 
Hint: See if the disparity is smaller for larger values of $p$. Also check for probabilities larger than 0.001. Recall we made an assumption when we calculated the interest rate.


```{r}
qnorm(0.001)
qnorm(0.002)
```


Your answer here.

## Problem 3

We were able to set an interest rate of about 2% that guaranteed a very low probability of having a loss. Furthermore, the expected average was over $1 million. Now other financial companies noticed the success of our business. They also noted that if we increase the number of loans we give, our profits increase. However, the pool of reliable borrowers was limited. So these other companies decided to give loans to less reliable borrowers but at a higher rate.

## Problem 3A

The pool of borrowers they found had a much higher default rate, estimated to be $p=0.05$. What interest rate would give these companies the same expected profit as your bank (Answer to 2E)? 

```{r}
#assuming the profit from 2E was 1 million dollars
p_default <- .05
total_profit <- 1000000

interest <- get_interest()
interest

#4%
```

## Problem 3B 

At the interest rate calculated in 3A what is the probability of negative profits? Use both the normal approximation and then confirm with a Monte Carlo simulation.

```{r}

mu <- total_profit
profit <- get_profit()

sigma <- sqrt(mean((profit-mu)^2))

#probibility that profit is less than 0
pr <- (pnorm((0-mu)/sigma))
pr


mean(profit<0)


```

## Problem 3C 

Note that the probability is much higher now. This is because the standard deviation grew. The companies giving out the loans did not want to raise interest rates much more since it would drive away clients. Instead they used a statistical approach. They increased $N$. How large does $N$ need to be for this probability to be 0.001? Use the central limit approximation and then confirm with a Monte Carlo simulation.

Your answer here.
```{r}



```

So by doubling the number of loans we were able to reduce our risk! Now, for this to work, all the assumptions in our model need to be approximately correct, including the assumption that the probability of default was **independent**. This turned out to be false and the main reason for the under estimation of risk.


## Problem 3D

Define the following matrix of outcomes for two borrowers using our previous box model:

```{r}
loan <- 180000
loss_per_foreclosure <- 120000
p2 <- 0.05
interest_rate2 <- 0.05
B <- 10^5
outcomes1 <- replicate(B,{
  sample( c(-loss_per_foreclosure, interest_rate2*loan ), 2, replace=TRUE, prob=c(p2, 1-p2))
})
```
We can confirm independence by computing the probability of default for the second conditioned on the first defaulting: 

```{r}
sum( outcomes1[1,] < 0 & outcomes1[2,]<0)/sum(outcomes1[1,]<0)
```

This quantity is about the same as the probability of default $0.05$.

Now we create a new model. Before generating each set of defaults, we assume that a random event occurred that makes all default probabilities go up or go down by 4 points. We could see how this would happen if, for example, demand for houses decreases and all house prices drop. 

```{r}
B <- 10^5
outcomes2 <- replicate(B,{
  add <- sample( c(-0.04,0.04), 1)
  sample( c(-loss_per_foreclosure, interest_rate2*loan ), 2, replace=TRUE, prob=c(p2+add, 1-(p2+add)))
})
```

Note that the outcomes are no longer independent as demonstrated by this result not being equal to 0.05

```{r}
sum( outcomes2[1,] < 0 & outcomes2[2,]<0)/sum(outcomes2[1,]<0)
```


Generate a simulation with correlated outcomes such as those above. This time use the interest rate calculated in 3A. What is the expected earnings under this model compared to the previous? What is the probability of losing $1 million compared to the previous? What is the probability of losing $10 million compared to the previous?



```{r}

B <- 10^5
outcomes2 <- replicate(B,{
add <- sample( c(-0.04,0.04), 1)
X<-sample( c(-loss_per_foreclosure, interest_rate2*loan ), 2, replace=TRUE, prob=c(p2+add, 1-(p2+add)))
sum(X)
})
outcomes2

```


## Problem 4

Read [this wikipedia page](https://en.wikipedia.org/wiki/Financial_crisis_of_2007%E2%80%9308) about the financial crisis. Write a paragraph describing how what you learned in this homework can help explain the conditions that led to the crisis.


